---
title: "Diet Breadth Model in foragerR"
author: "Michael Holton Price"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
The diet breadth model is a classic foraging model first introduced by MacArthur and Pianka (1966) in American Naturalist, "On Optimal Use of a Patchy Environment" (Volume 100, pages 603-609). This vignette describes the diet breadth model implemented in the open source R package forageR available at \url{https://github.com/MichaelHoltonPrice/forageR/}. The package can be installed with these commands:

```{r,message=F,eval=F}
library(devtools)
install_github('MichaelHoltonPrice/forageR')
```

## Model
A forager searches a landscape for resources indexed by $i$. Researches are searched for simultaneously and aach resource is encountered according to a Poisson process with encounter rate $\lambda_i$. When a resource is encountered, the forager chooses to procure it or ignore it and continue searching for other resources. If the forager procures it, the forager spends $h_i$ time procuring the resource (the handling time) and gains $v_i$ from procuring it (the value; usually energy in kilocalories).

Each resource has an on encounter return rate, or rank, given by

$$q_i = \frac{v_i}{h_i}$$

Let $b_i$ be a boolean indicator variable that is $1$ is a resource is procured on encounter and $0$ otherwise.
The overall return rate from foraging is

$$\dot{V} = \frac{\dot{c}_s + \sum_{i} \lambda_i \, v_i}{1 + \sum_{i} \lambda_i \, h_i} $$

where a dot above a variable indicates a time derivative and $\dot{c}_s$ is the rate at which value is lost during search (usually energy per time). There is a generally unique set of resources that one a forager should procure on encounter to maximize this overall return rate. This optimal set is found by sorting resources from best to worst according to the on encounter return rate $q_i$, then adding items to the diet so long as doing so improves the overall return rate. The first item excluded from the diet (and all remaining ones) has an on encounter return rate that is lower than the overall return rate achieved by procuring all higher ranked items.

## Example
The forageR package provides three functions related to the diet breadth model: `dietBreadthOptimalSet`, `dietBreadthReturnRate`, and `dietBreadthOptimalReturnRate`. `dietBreadthOptimalSet` determines the optimal set of resources to procure on encounter to maximize the overall return rate. `dietBreadthReturnRate` calculates the return rate for a set of resources, whether or not they are the optimal set. Optionally, items can be included in or excluded from the diet using the optional input variable `include`; if `include` is not specified, all items are used. `dietBreadthOptimalReturnRate` calculates the return rate for the optimal set. It is merely a wrapper function that calls `dietBreadthOptimalSet` to determine the optimal set then calls `dietBreadthReturnRate` with the variable `incldude` equal to the optimal set.

To demonstrate the use of these functions, load resource data from Winterhalder et al. (1988), "The Population Ecology of Hunter-Gatherers and Their Prey" (Journal of Anthropological Archaeology, Volume 7, Pages 289-328).

```{r}
library(forageR)
data(wintPrey)
print(wintPrey)
```

This data set contains eight notional resources named Aprey through Hprey. For each, the following is given: the energy value (kcal), handling time (min), procurement loss (kcal/min), density (No. / km^2), growth rate (1/year), and search radius (km). The search radius is the radius in around a forager in which the resource is visible. Winterhalder et al. (1988) assume a forager's search speed is 0.5 km / hr. The encounter rate, value accounting for procurement cost, and handling time are, respectively (ordered by rank),

```{r}
rank <- wintPrey$energy / wintPrey$handling_time
ord <- order(rank,decreasing=T)
wintPrey <- wintPrey[ord,]
forageSpeed <- 0.5 # km / hr
lambda <- wintPrey$density * 2 * wintPrey$search_radius * forageSpeed / 60 # No / min
v <- wintPrey$energy - wintPrey$procure_loss * wintPrey$handling_time # kcal
h <- wintPrey$handling_time # min
```

Calculate the return rate from progressively adding items to the diet in rank order, using a search cost of 4 kcal / min:

```{r}
include <- rep(F,length(lambda))
returnRate <- rep(NA,length(lambda))
search_cost <- 4 # kcal / min
for(i in 1:length(lambda)) {
  include[i] <- T
  returnRate[i] <- dietBreadthReturnRate(v,h,lambda,search_cost,include) # kcal / min
}
print(cbind(wintPrey$name,returnRate*60)) # kcal / hr
```

This reproduces the final column in Table 8 of Winterhalder et al. (1988). The optimal diet includes (in rank order) Cprey, Dprey, Aprey, and Bprey and excludes (in rank order) Fprey, Eprey, Hprey, and Gprey. This can be determined directly by calling `dietBreadthOptimalSet`.

```{r}
  optimalSet <- dietBreadthOptimalSet(v,h,lambda,search_cost)
  print(wintPrey$name[optimalSet])
```

The optimal return rate can be found by calling `dietBreadthReturnRate` with this optimal set or calling `dietBreadthOptimalReturnRate`, which first determines the optimal set then calculate the return rate:

```{r}
  returnRate1 <- dietBreadthReturnRate(v,h,lambda,search_cost,optimalSet)
  print(returnRate1*60) # kcal / hr
  returnRate2 <- dietBreadthOptimalReturnRate(v,h,lambda,search_cost)
  print(returnRate2*60) # kcal / hr
```
